### 2018-08-01 mlr_summary.tex
kernel trick

So the projection $z_i = V_P\T(x_i-\mu)$ is the projecting the \enquote{centered} $x_i$.
For the derivation of $V_P$ lets assume that $x_i$ are already centered ($E(x)=0$).
What we want to minimize is the projection error, which we can estimate by back projecting the projection of $x_i$ and comparing it with the initial $x_i$.
\eqnsnn {&
\hat{V}_P = \argmin_{V_P} \sum_{i=1}^N ||x_i- V_PV_P\T x_i||^2 
&&}
This time we will derive the optimum by observing the properties of the equation instead of root finding in the derivative.
Lets reformulate the loss a bit.
\eqnsnn {
L(V_P) &= \sum_{i=1}^N (x_i- V_PV_P\T x_i)\T(x_i- V_PV_P\T x_i)
&&\\
&= \sum_{i=1}^N x_i\T x_i - 2 x_i\T V_P V_P\T x_i 
+ x_i\T V_P V_P\T V_P V_P\T x_i
&&\\
& = \sum_{i=1}^N x_i\T x_i - 2x_i\T V_P V_P\T x_i  + x_i\T V_P V_P\T x_i
&&\\
&= \sum_{i=1}^N x_i\T x_i - x_i\T V_P V_P\T x_i
&&\\
&= C - \sum_{i=1}^N x_i\T V_P V_P\T x_i \with C = \sum_{i=1}^N x_i\T x_i
&&\\
&= C - \sum_{i=1}^N || V_P\T x_i ||^2 \mathtext{($V_P\T$ composed of row vectors $v_j\T$)}
&&\\
&= C - \sum_{i=1}^N \sum_{j=1}^P (v_j\T x_i)*(v_j\T x_i)
 \;=\; C - \sum_{j=1}^P \sum_{i=1}^N  (x_i\T v_j)*(x_i\T v_j)
&&\\
&= C - \sum_{j=1}^P ||X v_j||^2 \mathtext{($X$ composed of row vectors $x_i\T$)}
&&\\
&= C - \sum_{j=1}^P v_j\T X\T X v_j
&&}
So what we can see from this, is that the loss will be minimal if we maximize $\sum_{j=1}^P v_j\T X\T X v_j$.
We could again find the root of its derivative but, it will be more fun with some clever factorization of $X\T X$, which is the Eigendecomposition. 
Since $X\T X$ is a symmetric matrix, its Eigendecomposition is
\eqnsnn{&
X\T X = Q\Lambda Q\T \mathtext{with orthonormal} Q \mathtext{and diagonal} \Lambda
&&\\
& \argmax_{v_{1:P}} \sum_{j=1}^P v_j\T Q\Lambda Q\T v_j
&&} 
Suppose the eigenvalues $\lambda_1,\dots,\lambda_M$ are sorted in value descending order on the diagonal of $\Lambda$.
To maximize the above expression, we need to choose $v_{1:P}$ to be the first $P$ eigenvectors $q_i$ corresponding to the $P$ largest eigenvalues. 
If we could, we would choose all $v_j$ to be the first eigenvector, as this would truly yield the maximum, but due to our constraint of $v_j$s being orthonormal to each other we can't do that.

@@

>>>>

So the projection $z_i = V_P\T(x_i-\mu)$ is projecting the \enquote{centered} $x_i$.
For the derivation of $V_P$ lets assume that $x_i$ are already centered ($E(x)=0$).
What we want to minimize is the projection error, which we can estimate by back projecting the projection of $x_i$ and comparing it with the initial $x_i$.
\eqnsnn {&
\hat{V}_P = \argmin_{V_P} \sum_{i=1}^N ||x_i- V_PV_P\T x_i||^2 
&&}
This time we will derive the optimum by observing the properties of the equation instead of root finding in the derivative.
Lets reformulate the loss a bit.
\eqnsnn {
L(V_P) &= \sum_{i=1}^N (x_i- V_PV_P\T x_i)\T(x_i- V_PV_P\T x_i)
&&\\
&= \sum_{i=1}^N x_i\T x_i - 2 x_i\T V_P V_P\T x_i 
+ x_i\T V_P V_P\T V_P V_P\T x_i
&&\\
& = \sum_{i=1}^N x_i\T x_i - 2x_i\T V_P V_P\T x_i  + x_i\T V_P V_P\T x_i
&&\\
&= \sum_{i=1}^N x_i\T x_i - x_i\T V_P V_P\T x_i
&&\\
&= C - \sum_{i=1}^N x_i\T V_P V_P\T x_i \with C = \sum_{i=1}^N x_i\T x_i
&&\\
&= C - \sum_{i=1}^N || V_P\T x_i ||^2 \mathtext{($V_P\T$ composed of row vectors $v_j\T$)}
&&\\
&= C - \sum_{i=1}^N \sum_{j=1}^P (v_j\T x_i)*(v_j\T x_i)
 \;=\; C - \sum_{j=1}^P \sum_{i=1}^N  (x_i\T v_j)*(x_i\T v_j)
&&\\
&= C - \sum_{j=1}^P ||X v_j||^2 \mathtext{($X$ composed of row vectors $x_i\T$)}
&&\\
&= C - \sum_{j=1}^P v_j\T X\T X v_j
&&}
So what we can see from this, is that the loss will be minimal if we maximize $\sum_{j=1}^P v_j\T X\T X v_j$.
We could again find the root of its derivative but, it will be more fun with some clever factorization of $X\T X$, which is the Eigendecomposition. 
Since $X\T X$ is a symmetric matrix, its Eigendecomposition is
\eqnsnn{&
X\T X = Q\Lambda Q\T \mathtext{with orthonormal} Q \mathtext{and diagonal} \Lambda
&&\\
& \argmax_{v_{1:P}} \sum_{j=1}^P v_j\T Q\Lambda Q\T v_j
&&} 
Suppose the eigenvalues $\lambda_1,\dots,\lambda_M$ are sorted in value descending order on the diagonal of $\Lambda$.
To maximize the above expression, we need to choose $v_{1:P}$ to be the first $P$ eigenvectors $q_i$ corresponding to the $P$ largest eigenvalues. 
If we could, we would choose all $v_j$ to be the first eigenvector, as this would truly yield the maximum, but due to our constraint of $v_j$s being orthonormal to each other we can't do that.

@@

\section{The Kernel Trick}
Remember when we introduced features to make things non-linear in linear regression.
We use features to calculate new quantities from a given datapoint such as 
\\$\phi(x) = \mat{1 & x & x^2 & e^x & e^{-x} & \sin(x) & \cos(x)}\T$.
This is a very powerful way to allow for non-linearity but on the downside of it is the question:
How to choose features so that some property of the original function or distribution becomes linear?
This is not easy and sometimes coming up with the right features is impossible.

The kernel trick is a way to get rid of explicitly defining features.
Before diggin in, lets define what a kernel in this sense even is.
A kernel is the dot product of the feature vectors of two inputs $x$ and $x'$.
\eqnsnn{&
k(x,x') = \phi(x)\T\phi(x')
&&}
This implies that a kernel is symmtric with respect to argument order $k(x,x') = k(x',x)$.
Of course when we difine features, we imply a kernel function.
If on the other hand we define a kernel function, we imply features.
Due to the kernel being a dot product, it is a measure of similarity of the feature vectors and thus also a measure of similarity of $x$ and $x'$.
If we keep this in mind, we can define kernels with respect to the distance of two points, for example a gaussian kernel:
\eqnsnn{&
k_\mathcal{N}(x,x') = \mathcal{N}(||x-x'||, \mu, \sigma^2) = \frac1{\sqrt{2\pi\sigma^2}}e^{\frac{(||x-x'||-\mu)^2}{2\sigma^2}}
&&}

The \enquote{trick} in the kernel trick is now to get rid of the features $\phi(x)$ by reformulating the problem to only include dot products $\phi(x)\T\phi(x')$, which are then replaced by kernels $k(x,x')$.
Lets illustrate this process on various examples in the following.

\subsection{Kernel PCA}
In PCA we have following eigenvalue problem.
\eqnsnn{&
X\T X  v_j = \lambda v_j \mathtext{where $v_j$ is one of the eigenvectors used for the projection}
&&\\
&
\lambda v_j = X\T X v_j = \sum_{i=1}^N \phi(x_i) \phi(x_i)\T v_j \mathtext{(sum of outer products of $\phi(x_i)$)}
&&}
From this we can extract a calculation for $v_j$ by replacing the dot product of $v_j$ and $\phi(x_i)$ by a new variable $b_{ij}$:
\eqnsnn{&
\sum_{i=1}^N \phi(x_i) \phi(x_i)\T v_j = \sum_{i=1}^N \phi(x_i) b_{ij} = \lambda v_j
&&\\&
\sum_{i=1}^N \phi(x_i) \alpha_{ij} = v_j \with \alpha_{ij} = \frac{b_{ij}}{\lambda}
&&\\&
\sum_{i=1}^N \phi(x_i) \alpha_{ij} = X\T\alpha_j = v_j
&&} 
When we substitue this into the initial eigenvalue problem we get the following we get to the following.
\eqnsnn{
X\T X  v_j = \lambda v_j \quad&=\quad X\T X X\T \alpha_j = \lambda X\T\alpha_j
&&\\
&\equivalent\quad XX\T XX\T \alpha_j = \lambda XX\T\alpha_j
&&}
This is pretty much what we have been waiting for. 
We generated an expression using matrices of dot products of features, which we can replace by a kernel matrix \\$XX\T_{ik} = \phi(x_i)\T\phi(x_k) = k(x_i,x_k) = K_{ik}$.
This brings us to a new eigenvalue problem which only contains kernel expressions.
\eqnsnn{&
K K \alpha_j = \lambda K\alpha_j \quad\equivalent\quad K\alpha_j = \lambda\alpha_j
&&}

We probably can get to this a little bit more straight forward since we know we want $XX\T$ in our equation:
\eqnsnn{
X\T X  v_j = \lambda v_j \quad\equivalent\quad X\T XX\T  v_j &= \lambda X\T v_j 
&&\\ \quad\equivalent\quad XX\T XX\T  v_j &= \lambda XX\T v_j
&&\\ \quad\equivalent\quad K v_j &= \lambda v_j
&&}
